---
title: "Real-Data Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-Data Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`ggSynteny` currently supports two pairwise input routes:

- `MCScanX`: `gff + collinearity`
- `jcvi`: `BED + anchors`

Both routes produce the same internal `ggsynteny_data` object, so the plotting
and annotation steps are shared.

This vignette is written as a real-data template. The examples below use the
bundled toy files so they can run during package checks, but the same code
structure is intended for real genome pairs.

# Data preparation checklist

Before running a real dataset, check these points:

1. Gene IDs in the pairwise synteny output and annotation files are consistent.
2. Chromosome names are consistent between pairwise input and `GFF3`.
3. Coordinates are all on the same genome build.
4. The pairwise result contains enough anchors per block for the plot you want.
5. For transcript-aware micro-synteny, the `GFF3` actually includes transcript,
   exon, or CDS features.

If gene IDs are not identical between the pairwise input and `GFF3`, prepare a
named mapping vector and pass it through `gene_id_map1` and `gene_id_map2`.

# MCScanX workflow

```{r}
library(ggSynteny)

mcscanx_gff <- system.file("extdata", "toy_mcscanx.gff", package = "ggSynteny")
collinearity <- system.file("extdata", "toy_mcscanx.collinearity", package = "ggSynteny")
annotation <- system.file("extdata", "toy_annotation.gff3", package = "ggSynteny")

res_mcscanx <- validate_real_synteny_workflow(
  mcscanx_gff = mcscanx_gff,
  collinearity = collinearity,
  genome1 = "GenomeA",
  genome2 = "GenomeB",
  gff3_genome1 = annotation,
  gff3_genome2 = annotation,
  transcript_mode = "canonical",
  min_pairs = 2,
  flank_genes = 1
)
```

The returned object contains:

```{r}
names(res_mcscanx)
names(res_mcscanx$data)
```

Genome-level and micro-synteny plots are both created from the same validated
data object:

```{r, fig.height=4, fig.width=7}
res_mcscanx$genome_plot
```

```{r, fig.height=4, fig.width=7}
res_mcscanx$micro_plot
```

# jcvi workflow

```{r}
jcvi_anchors <- system.file("extdata", "toy_jcvi.anchors", package = "ggSynteny")
bed_a <- system.file("extdata", "toy_jcvi_genomeA.bed", package = "ggSynteny")
bed_b <- system.file("extdata", "toy_jcvi_genomeB.bed", package = "ggSynteny")

res_jcvi <- validate_jcvi_synteny_workflow(
  anchors = jcvi_anchors,
  bed_a = bed_a,
  bed_b = bed_b,
  genome1 = "GenomeA",
  genome2 = "GenomeB",
  gff3_genome1 = annotation,
  gff3_genome2 = annotation,
  transcript_mode = "canonical",
  min_pairs = 2,
  flank_genes = 1
)
```

```{r, fig.height=4, fig.width=7}
res_jcvi$genome_plot
```

```{r, fig.height=4, fig.width=7}
res_jcvi$micro_plot
```

# Real-data substitution points

When replacing the toy files with real data, these are the lines you usually
change:

```r
res <- validate_real_synteny_workflow(
  mcscanx_gff = "species_pair.gff",
  collinearity = "species_pair.collinearity",
  genome1 = "SpeciesA",
  genome2 = "SpeciesB",
  gff3_genome1 = "speciesA.gff3",
  gff3_genome2 = "speciesB.gff3",
  gene_id_map1 = NULL,
  gene_id_map2 = NULL,
  transcript_mode = "canonical",
  min_pairs = 5,
  flank_genes = 3
)
```

Or, for `jcvi`:

```r
res <- validate_jcvi_synteny_workflow(
  anchors = "species_pair.anchors",
  bed_a = "speciesA.bed",
  bed_b = "speciesB.bed",
  genome1 = "SpeciesA",
  genome2 = "SpeciesB",
  gff3_genome1 = "speciesA.gff3",
  gff3_genome2 = "speciesB.gff3",
  gene_id_map1 = NULL,
  gene_id_map2 = NULL,
  transcript_mode = "canonical",
  min_pairs = 5,
  flank_genes = 3
)
```

# Troubleshooting

- No blocks parsed: confirm the pairwise input really contains anchors for the
  selected genome pair.
- Empty feature layer: confirm the `GFF3` contains transcript, exon, or CDS
  entries and that `gene_id_map*` is correct.
- Missing strand or labels: confirm those values exist in the `GFF3` and match
  the same gene IDs used in the pairwise output.
- Micro-synteny looks too sparse: reduce `min_pairs`, reduce `flank_genes`, or
  choose a block with more anchors.
